---
title: "Power Calculations"
author: "CT Berezin"
date: "11/19/2021"
output: html_document
---

Note: SD is the square root of the variance (so square SD to find variance)

Is MOR circadian/light-driven? (qPCR)
SD within groups is usually around 0.2
Same as MOR changes in morphine vs saline treatment
```{r}
power.anova.test(groups = 4, within.var = 0.04,
                 between.var = 0.04,
                 sig.level = 0.05, power = 0.9)
```

Is MOR circadian/light-driven? (Western blot)
Quantification of the band density
20% variation within groups
30% significant change
Same as in morphine vs saline groups
```{r}
power.anova.test(groups = 4, within.var = 0.04,
                 between.var = 0.09,
                 sig.level = 0.05, power = 0.9)
```

Is MOR function circadian/light-driven? (GTPyS)
Results are typically presented as a %
A 10% SD within groups
A 20% change being significant
```{r}
power.anova.test(groups = 4, within.var = 0.01,
                 between.var = 0.04,
                 sig.level = 0.05, power = 0.9)
```


Morphine + B-endorphin changes
Same as morphine + MOR changes
***check out values from existing data
```{r}
power.anova.test(groups = 8, within.var = ,
                 between.var = ,
                 sig.level = 0.05, power = 0.8)
```




Methylation

```{r}
power.anova.test(groups = 4, within.var = ,
                 between.var = ,
                 sig.level = 0.05, power = 0.8)
```
